Wastewater Surveillance for SARS-CoV-2 Data Analysis Project

### 2023-04-25 code/modeling_code/2_log10VL_7DMA_model.Rmd
update manuscript

```{r}
recipe_7dma <- recipe(pcr_pos_7dma ~ log10_VL, data = train_data)
```

@@

```{r}
aug_test <- augment(fit_7dma, train_data)
aug_test %>% select(pcr_pos_7dma, .pred)
```

@@

```{r}
rmse <- aug_test %>% rmse(truth = pcr_pos_7dma, .pred)
rsq <- aug_test %>% rsq(truth = pcr_pos_7dma, .pred)
log10VL_7DMA_metrics <- full_join(rmse, rsq)

@@

```{r}
aug_test %>% ggplot(aes(pcr_pos_7dma, .pred)) +
  geom_point() +
  stat_smooth(method = "lm")
ggsave(here("data/modeling_data/metrics/log10VL_7DMA_predict.png"))
```

@@

```{r}
aug_residual <- aug_test %>% mutate(residual = pcr_pos_7dma - .pred)
aug_residual %>% ggplot(aes(.pred, residual)) +
  geom_point()

>>>>

```{r}
recipe_7dma <- recipe(case_7dma ~ log10_VL, data = train_data)
```

@@

```{r}
aug_test <- augment(fit_7dma, train_data)
aug_test %>% select(case_7dma, .pred)
```

@@

```{r}
rmse <- aug_test %>% rmse(truth = case_7dma, .pred)
rsq <- aug_test %>% rsq(truth = case_7dma, .pred)
log10VL_7DMA_metrics <- full_join(rmse, rsq)

@@

```{r}
aug_test %>% ggplot(aes(case_7dma, .pred)) +
  geom_point() +
  stat_smooth(method = "lm")
ggsave(here("data/modeling_data/metrics/log10VL_7DMA_predict.png"))
```

@@

```{r}
aug_residual <- aug_test %>% mutate(residual = case_7dma - .pred)
aug_residual %>% ggplot(aes(.pred, residual)) +
  geom_point()

### 2023-04-25 code/modeling_code/4_AssayPos_7DMA_model.Rmd
update manuscript

---
title: "PP_7DMA_model"
author: "Leah Lariscy"
date: "2023-03-17"
output: html_document
---

@@

Predictor: Percent positive technical replicates per sample

@@

This is where we will set PP as predictor and 7DMA as outcome

@@

```{r}
rmse <- aug_test %>% rmse(truth = case_7dma, .pred)
rsq <- aug_test %>% rsq(truth = case_7dma, .pred)
huber <- aug_test %>% huber_loss(truth = case_7dma, .pred)
PP_7DMA_metrics <- bind_rows(rmse, rsq, huber)

saveRDS(PP_7DMA_metrics, here("data/modeling_data/metrics/PP_7DMA_performance.RDS"))

PP_7DMA_metrics

@@

```{r}
aug_test %>% ggplot(aes(case_7dma, .pred)) +
  geom_point() +
  stat_smooth(method = "lm")
ggsave(here("data/modeling_data/metrics/PP_7DMA_predict.png"))
```

@@

ggsave(here("data/modeling_data/metrics/PP_7DMA_residual.png"))
```

## Plot 7DMA and prediction temporally

```{r}
aug_test %>% ggplot(aes(x=date)) +
  geom_point(aes(y=case_7dma), color = "blue") +
  geom_line(aes(y=case_7dma), color = "blue") +
  geom_point(aes(y=.pred), color = "green") +
  geom_line(aes(y=.pred), color = "green")

## Use model on testing data

```{r}
aug_test2 <- augment(fit_7dma, test_data)
aug_test2 %>% select(case_7dma, .pred)
```

```{r}
aug_test2 %>% ggplot(aes(x=date)) +
  geom_point(aes(y=case_7dma), color = "blue") +
  geom_line(aes(y=case_7dma), color = "blue") +
 geom_point(aes(y=.pred), color = "green") +
  geom_line(aes(y=.pred), color = "green")
```

## Evaluate model performance on testing set

```{r}
rmse <- aug_test2 %>% rmse(truth = case_7dma, .pred)
rsq <- aug_test2 %>% rsq(truth = case_7dma, .pred)
huber <- aug_test2 %>% huber_loss(truth = case_7dma, .pred)
PP_7DMA_test_metrics <- bind_rows(rmse, rsq, huber)

saveRDS(PP_7DMA_test_metrics, here("data/modeling_data/metrics/PP_7DMA_test_performance.RDS"))

PP_7DMA_test_metrics
```

## Use model on testing dates

```{r}
aug_test3 <- augment(fit_7dma, test_dates)
aug_test3 %>% select(case_7dma, .pred)
```

## Evaluate model performance on testing dates

```{r}
rmse <- aug_test3 %>% rmse(truth = case_7dma, .pred)
rsq <- aug_test3 %>% rsq(truth = case_7dma, .pred)
huber <- aug_test3 %>% huber_loss(truth = case_7dma, .pred)
PP_7DMA_BA5_metrics <- bind_rows(rmse, rsq, huber)

saveRDS(PP_7DMA_BA5_metrics, here("data/modeling_data/metrics/PP_7DMA_BA5_performance.RDS"))

PP_7DMA_BA5_metrics
```

```{r}
aug_test3 %>% ggplot(aes(x=date)) +
  geom_point(aes(y=case_7dma), color = "blue") +
  geom_line(aes(y=case_7dma), color = "blue") +
 geom_point(aes(y=.pred), color = "green") +
  geom_line(aes(y=.pred), color = "green")
```

>>>>

---
title: "assaypos_7dma_model"
author: "Leah Lariscy"
date: "2023-03-17"
output: html_document
---

@@

Predictor: qPCR assay positivity

@@

This is where we will set assaypos as predictor and 7dma as outcome

@@

```{r}
rmse <- aug_test %>% rmse(truth = case_7dma, .pred)
rsq <- aug_test %>% rsq(truth = case_7dma, .pred)
assaypos_7dma_metrics <- full_join(rmse, rsq)

saveRDS(assaypos_7dma_metrics, here("data/modeling_data/metrics/assaypos_7dma_performance.RDS"))

assaypos_7dma_metrics

@@

```{r}
aug_test %>% ggplot(aes(case_7dma, .pred)) +
  geom_point() +
  stat_smooth(method = "lm")
ggsave(here("data/modeling_data/metrics/assaypos_7dma_predict.png"))
```

@@

ggsave(here("data/modeling_data/metrics/assaypos_7dma_residual.png"))
```

## Final fit on test data, then on test dates

```{r}
#testing data
fit_test_data <- workflow_7dma %>% 
  fit(data = test_data)

aug_test <- augment(fit_test_data, test_data)
aug_test %>% select(case_7dma, .pred)

rmse <- aug_test %>% rmse(truth = case_7dma, .pred)
rsq <- aug_test %>% rsq(truth = case_7dma, .pred)
test_metrics <- full_join(rmse, rsq)

test_metrics
```

```{r}
#testing dates
fit_test_dates <- workflow_7dma %>% 
  fit(data = test_dates)

aug_test_dates <- augment(fit_test_dates, test_dates)
aug_test_dates %>% select(case_7dma, .pred)

rmse <- aug_test_dates %>% rmse(truth = case_7dma, .pred)
rsq <- aug_test_dates %>% rsq(truth = case_7dma, .pred)
test_dates_metrics <- full_join(rmse, rsq)

test_dates_metrics
```

## Plot testing dates predictions vs truth

```{r}
aug_test_dates %>% ggplot(aes(x=date)) +
  stat_smooth(aes(y=case_7dma), method = "loess", span = 0.5, color = "black") +
  stat_smooth(aes(y=.pred), method = "loess", span = 0.5)
```

### 2023-04-25 code/modeling_code/5_AssayPos_TestPos_model.Rmd
update manuscript

---
title: "log10VL_testpos_model"
author: "Leah Lariscy"
date: "2023-03-17"
output: html_document
---

@@

Predictor: log10 wastewater viral load

Outcome: Clinical case seven day moving average

@@

This is where we will set log10VL as predictor and testpos as outcome

```{r}
recipe_testpos <- recipe(percent_pos_test ~ log10_VL, data = train_data)
```

@@

```{r}
workflow_testpos <- workflow() %>% 
  add_model(lm_mod) %>% 
  add_recipe(recipe_testpos)
```

@@

```{r}
set.seed(626)
fit_testpos <- workflow_testpos %>% 
  fit(data = train_data)

tidy(fit_testpos)
```

@@

```{r}
aug_test <- augment(fit_testpos, train_data)
aug_test %>% select(percent_pos_test, .pred)
```

@@

```{r}
rmse <- aug_test %>% rmse(truth = percent_pos_test, .pred)
rsq <- aug_test %>% rsq(truth = percent_pos_test, .pred)
log10VL_testpos_metrics <- full_join(rmse, rsq)

saveRDS(log10VL_testpos_metrics, here("data/modeling_data/metrics/log10VL_testpos_performance.RDS"))

log10VL_testpos_metrics

@@

```{r}
aug_test %>% ggplot(aes(percent_pos_test, .pred)) +
  geom_point() +
  stat_smooth(method = "lm")
ggsave(here("data/modeling_data/metrics/log10VL_testpos_predict.png"))
```

@@

ggsave(here("data/modeling_data/metrics/log10VL_testpos_residual.png"))
```

@@

```{r}
#testing data
fit_test_data <- workflow_testpos %>% 
  fit(data = test_data)

@@

```{r}
#testing dates
fit_test_dates <- workflow_testpos %>% 
  fit(data = test_dates)

>>>>

---
title: "assaypos_7dma_model"
author: "Leah Lariscy"
date: "2023-03-17"
output: html_document
---

@@

Predictor: qPCR assay positivity

Outcome: PCR test positivity rate (daily)

@@

This is where we will set assaypos as predictor and 7dma as outcome

```{r}
recipe_7dma <- recipe(percent_pos_test ~ pos_rate_tr, data = train_data)
```

@@

```{r}
workflow_7dma <- workflow() %>% 
  add_model(lm_mod) %>% 
  add_recipe(recipe_7dma)
```

@@

```{r}
set.seed(626)
fit_7dma <- workflow_7dma %>% 
  fit(data = train_data)

tidy(fit_7dma)
```

@@

```{r}
aug_test <- augment(fit_7dma, train_data)
aug_test %>% select(percent_pos_test, .pred)
```

@@

```{r}
rmse <- aug_test %>% rmse(truth = percent_pos_test, .pred)
rsq <- aug_test %>% rsq(truth = percent_pos_test, .pred)
assaypos_7dma_metrics <- full_join(rmse, rsq)

saveRDS(assaypos_7dma_metrics, here("data/modeling_data/metrics/assaypos_7dma_performance.RDS"))

assaypos_7dma_metrics

@@

```{r}
aug_test %>% ggplot(aes(percent_pos_test, .pred)) +
  geom_point() +
  stat_smooth(method = "lm")
ggsave(here("data/modeling_data/metrics/assaypos_7dma_predict.png"))
```

@@

ggsave(here("data/modeling_data/metrics/assaypos_7dma_residual.png"))
```

@@

```{r}
#testing data
fit_test_data <- workflow_7dma %>% 
  fit(data = test_data)

@@

```{r}
#testing dates
fit_test_dates <- workflow_7dma %>% 
  fit(data = test_dates)

### 2023-04-25 code/processing_code/processing.qmd
update manuscript

table1::label(final_data_sum$log10_VL) <- "Log10 Viral Load"
table1::label(final_data_sum$pos_rate_tr) <- "Assay Positivity"
table1::label(final_data_sum$case_7dma) <- "Clinical Case Seven-Day Moving Average"
table1::label(final_data_sum$pcr_pos) <- "Daily Reported Clinical Cases"
table1::label(final_data_sum$percent_pos_test) <- "PCR Test Positivity"

>>>>

#table1::label(final_data_sum$log10_VL) <- "Log10 Viral Load"
#table1::label(final_data_sum$pos_rate_tr) <- "Assay Positivity"
#table1::label(final_data_sum$case_7dma) <- "Clinical Case Seven-Day Moving Average"
#table1::label(final_data_sum$pcr_pos) <- "Daily Reported Clinical Cases"
#table1::label(final_data_sum$percent_pos_test) <- "PCR Test Positivity"

### 2023-04-25 products/manuscript/Manuscript.qmd
update manuscript

### Feature selection

Two wastewater and two clinical case metrics were identified as features of interest during exploratory analysis steps where

### Data splitting

Compare relationship differences between year 1 and year 2

### Model training and fitting

Trained linear regression models on year 1 data using cross validation

Performance validation, metrics?

### Model testing

The trained model was then tested on the year 2 data

@@

@@

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
#resulttable=readRDS("../../results/summarytable.rds")
#knitr::kable(resulttable)
```

![Distribution of recorded seven-day moving average for each day](images/7DMA_histo.png)

![Distribution of viral load per sample collection per target](images/total_copies_histo.png)

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
#knitr::include_graphics(here("results","height_weight_stratified.png"))
```

![Clinical case seven-day moving average over time](images/7DMA_time_plot.png)

![Wastewater viral load over time](images/avg_total_copies_time.png)

## Basic statistical analysis

![Linear regression of viral load vs case seven-day moving average](images/total_copies_avg_vs_7DMA.png)

## Full analysis

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
#resulttable2 = readRDS(here("results","resulttable2.rds"))
#knitr::kable(resulttable2)
```

>>>>

Through exploration of time series and wastewater-clinical correlations, prominent relationships were identified for further analysis. Wastewater variables were used to train linear regression models to predict clinical observations. Model performance metrics (RSQ and RMSE) were assessed against each other and against a null model. Cross-validation was used to validate performances. The best model was then determined and used to predict values in a new set of data.

@@

## Time-series analysis

## Correlation analysis

Two wastewater and two clinical case metrics were identified as features of interest during exploratory analysis steps where stronger wastewater-clinical correlations were observed between these variables than in others. These include wastewater viral load, wastewater assay positivity rate, clinical case seven-day moving average, and test administration positivity rate (Table 1).

@@

## Model analysis

```{r, echo=FALSE, fig.cap="Observed Clinical Cases vs Wastewater Predictions"}
include_graphics(here("figures/modeling/assaypos_7dma.png"))
```
